Modelling BCG vaccination in the UK

### 2019-08-02 rmds/08-model-development.Rmd
ellen simple model dev commments

Non-UK born pulmonary cases was estimated using the ETS (see Chapter \@ref(data)) for each age-group included in the model from 2000 until 2015. Prior to 2000, incidence in the non-UK born are unavailable and the relationship to transmission in UK-born cases is unknown. To account for this, importation of non-UK born cases in the model begins in 1960 and then is scaled up through to 2000. As the form of this relationship is unknown the following functional form was used to scale cases based on those observed in 2000,

\begin{equation}
\text{Non UK-born scaling} = \frac{e^{\left(\frac{t - 1960}{\text{ln}(2) * (-1) * \iota_{\text{scale}}}\right)} - 1}{e^{\left(\frac{2000 - 1960}{\text{ln}(2) * (-1) * \iota_{\text{scale}}}\right)} - 1}
  (\#eq:non-uk-born-scale)
\end{equation}

@@

The probability of transmission can be defined as the probability that a single contact between an infectious active TB case and a susceptible individual will lead to TB infection. The probability of transmission ($\beta_a$) can be redefined in terms of effective contacts ($C_{\text{eff}}$), historic effective contacts ($C^{\text{hist}}_{\text{eff}}$), actual average yearly total contacts ($C_{\text{actual}}$), the average period of time infectious ($\frac{1}{\nu_{\text{avg}}^{P}}$), and the average mortality rate ($\mu_{\text{age}}$) as follows,

@@

Vynnycky et al. found that the effective contact rate (i.e the transmission probability times the contact rate) for TB was approximately 22 in 1900 and fell to approximately 1 in 1990.[@Vynnycky1999] Incidence rates have increased since the early 1980's (Chapter \@ref(data)) and it is unclear what impact this has had on the effective contact rate. I have assumed that the effective contact rate is normally distributed with a mean of 1 and a standard deviation of 0.5. For the historic effective contact rate I have assumed a uniform distribution with a lower bound of the current effective contact rate and an upper bound of 20. I have also assumed that the historic contact rate declines over time, starting in 1935, reducing to the current effective contact rate in 1980. The speed of this decay is set by defining a half-life ($C^{\text{half-life}}_{\text{eff}}$). The prior for the half-life is assumed to be normally distributed with a mean of 5 years and a standard deviation of 5 (truncated to be greater than 0). This prior is based on the observed trend in notifications. Additional age stratification of $\beta$ is explored by including modifiers for certain age-groups. The baseline scenario is that no modification is required, with variation explored for young adults (15-29; $\beta_{\text{young adult}}$) as a scenario. The priors for this modifier was assumed to be uniform, bounded by 0 and 10. The contact rate is estimated by averaging the total age-specific contact rates estimated from POLYMOD data (Section \@ref(contact-matrix)) on an annual basis.

@@

```{r tb-time-treat, fig.scap = "Distribution of time to treatment (days) from the date of reported symptom onset until the date started treatment for the UK born, stratified by age group and pulmonary/extra-pulmonary TB status in the ETS system for notifications between 2000 and 2012.", fig.cap = "Distribution of time to treatment (days) from the date of reported symptom onset until the date started treatment for the UK born, stratified by age group and pulmonary/extra-pulmonary TB status in the ETS system for notifications between 2000 and 2012. Age is stratified into three groups; children (0-14), adults (15-69) older adults (70-89). The time from symptom onset to starting treatment is shorter for cases with pulmonary TB cases across age groups, with younger cases starting treatment more rapidly than older cases.", out.extra = ""}
knitr::include_graphics(file.path(resources_path, "resources", "figure", "time_to_treatment.png"))
```

@@

```{r tb-treat-time-succ, fig.scap = "Distribution of time to treatment completion in the UK born successfully treated (days), stratified by age group and pulmonary/extra-pulmonary TB status in the ETS for notifications between 2000 and 2012.", fig.cap = "Distribution of time to treatment completion in the UK born successfully treated (days), stratified by age group and pulmonary/extra-pulmonary TB status in the ETS for notifications between 2000 and 2012. Age is stratified into three groups;children (0-14), adults (15-69) older adults (70-89). There is little evidence that the time to successful treatment differs between pulmonary and extra-pulmonary cases only but older cases appear to have a high likelihood of longer treatment times.", out.extra = ""}
knitr::include_graphics(file.path(resources_path, "resources", "figure", "time_cure.png"))
```

@@

In the model presented here unstratified social (nonphysical) contacts are used to generate an age-stratified contact matrix. There are several reasons for this. Firstly, stratifying by home, school, work, transport or leisure contacts whilst initially appealing as doing so may lead to insights as to the nature of the type of contacts required for TB transmission may also lead to over-fitting without a strong apriori hypothesis. In high and medium burden countries it has been shown that within household transmission is not a major driver of overall transmission.[@Mathema2018] Until recently it has been thought that household transmission plays a more dominated role in low burden settings, such as England, which would indicate that home contacts should be considered. However, it has recently been found that only 7.7% (1849/24,060) of cases in England between 2010 and 2012 lived in a household with another case.[@Lalor2017] The same study estimated that overall only 3.9% of cases were due to recent household transmission, and there was no evidence that cases within households were more likely to transmit within the household than outside of it. There is little evidence to suggest that school, work, transport or leisure contacts are more likely to transmit TB in England than any other contact. The choice of contact type is disease dependent; for TB it is likely that closer contacts result in a greater likelihood of transmission.[@Mathema2018] Unfortunately the physical contacts recorded in the POLYMOD survey represent a poor proxy to closeness of contacts as physical contact can be a little as a handshake and because TB is a respiratory disease physical contact is not required. For this reason physical contacts have not been further evaluated. Instead, the uncertainty in age-dependent transmission rates has been explored by allowing for scenarios in which the transmission probability varies between children and adults, between children, adults and older adults and in which it does not vary.

@@

As the POLYMOD contact data was collected using a survey there is likely to be measurement error and missing data for the number of contacts reported and the age that contacts were reported to be. Some participants also recorded contacts with an estimated age range rather than with a point estimate. In addition, as the survey had a relatively low sample size (1,011) in the UK the estimated contact matrices contain considerable uncertainty. These considerations are often not considered in modelling studies but may introduce significant bias. Here the `socialmixr` R package is used to generate 1000 bootstrapped contact matrix samples using the following steps,

@@

Where $N_i$ is the number of people in age group $i$, $N_j$ is the number of people in age group $j$, $c_{ij}$ is the number of contacts between members of group $i$ with group $j$ and $c_{ji}$ is the number of contacts between members of group $j$ with group $i$. In the POLYMOD survey this relationship does not hold exactly due to random variation. A symmetric contact matrix ($C^{\prime}_{ij}$) can be derived by averaging the contacts between the $i$ and $j$ groups and the $j$ and $i$ groups for all age groups using the following equation,

\begin{equation} 
  C^{\prime}_{ij} = \frac{C_{ij}N_i + C_{ji}N_j}{2N_i}
  (\#eq:symmfull)
\end{equation}

@@

```{r contact-tile-plot, out.width = "250px", fig.scap = "a.) Mean contact rates and the b.) normalised standard deviation (\\%) of 1000 boostrapped samples of  social contacts from the POLYMOD social contact survey using 5 year age groups up 49 years old and then a single group for 50-69 year olds.", fig.cap = "a.) Mean contact rates and the b.) normalised standard deviation (\\%) of 1000 boostrapped samples of  social contacts from the POLYMOD social contact survey using 5 year age groups up 49 years old and then a single group for 50-69 year olds. Mixing is highly assortative by age with children and young adults representing the majority of contacts. There is also evidence of mixing between children and middle age adults with older children mixing with progressivly older adults. Contact rates in older adults are highly uncertain, with the most uncertainty in mixing between older adults and young children."}
knitr::include_graphics(file.path(resources_path, "resources", "figure", "contact_matrix.png"))
```

@@

An overview of the vacination model parameters can be found in: Table \@ref(tab:disease-model) for parameters that impact the natural history of TB; Table  \@ref(tab:demo-model-params) for parameters that impact the population level distribution of BCG vaccination; and Table \@ref(tab:sources-tab) for an overview of the sources used to generate prior distributions. More detail is given, where required, in the following section.

@@

The effectiveness of the BCG vaccine is usually estimated using it's effectiveness at reducing the incidence of active TB cases in a susceptible population. In the model outlined in this chapter the action of the BCG vaccine has been split into its main effect of reducing the rate of latent TB cases developing active disease and its secondary effect of reducing the likelihood of initial infection. There are few estimates of the effectiveness of the BCG vaccine at preventing active TB in cases that are already latently infected and where these estimates do exist they are not stratified by time since vaccination, or age at vaccination.[@Roy2014] The overall effectiveness ($\alpha^T_a$) of the BCG vaccine can be estimated from the combined effectiveness at preventing initial infection ($\chi_a$) and then preventing activation in latently infected individuals ($\alpha_a$) using the following equation,

@@

```{r bcg-eff-est}
kable(eff_bcg, booktabs = TRUE, escape = TRUE,
      caption.short = "Estimates of the effectiveness of the BCG vaccine at preventing active TB disease stratified by years since vaccination.", 
      caption = "Estimates of the effectiveness of the BCG vaccine at preventing active TB disease stratified by years since vaccination. For 0-9 years since vaccination estimates were derived using Poisson regression from the MRC BCG trial and for 10-29 years since vaccination estimates were extracted from a more recent case control cohort study in the UK born vaccinated at school-age.[@Hart1972; @Mangtani2017] BCG effectiveness at preventing active TB disase used as prior distributions in the model.") %>% 
  kable_styling(latex_options = "hold_position")
```

@@

\begin{equation}
  \alpha^T_a = 1 - e^{\mathcal{N}(\text{ln}(RR), SE)}
  (\#eq:rr-to-alpha)
\end{equation}

Where $SE$ is the standard error of the logged Risk Ratio. The transformed values used as the prior distribution are detailed in Table \@ref(tab:disease-model).

@@

Roy et al. published a meta-analysis that estimated the effectiveness of the BCG vaccine at preventing initial infection in children.[@Roy2014] This has been used as the primary source for this parameter, with the assumption being made that the effectiveness is the same in adults as it is in children. This is reasonable to assume as there is little evidence the the overall effectiveness of the BCG vaccines reduces with the age it is given in England. Unfortunately the meta-analysis by Roy et al. did not include an estimate of the effectiveness of the BCG vaccine at preventing initial TB infection stratified by time since vaccination. This is problematic as there is a large amount of evidence that the overall effectiveness of the BCG vaccine wanes with time,[@Abubakar2013; @Mangtani2017] and if the protection from initial infection does not also reduce over time then as overall effectiveness decreases the contribution from the prevention of initial infection will increase. For this reason I have assumed that the protection from initial infection reduces over time with the same functional form as for the overall effectiveness of BCG vaccination. This relation can be formalised using the following equation,

@@

Age-stratified and UK birth stratified population estimates for England were estimated using the LFS (Chapter \@ref(data)). Figure \@ref(fig:age-strat-demo) indicates that the age distribution of the UK born population changed over the study period (2000 to 2015), with an increase in those in late middle age and older and a decrease in those in early middle age. The proportion of young adults and young children also increased. This may have impacted TB incidence as young adults are thought to be responsible for the majority of transmission. Data from the 1931 census was also used to estimate the population of England in 1931 stratified into the modeled age groups.

@@

The time varying, age-specific, noisy, all-cause mortality rates ($\mu^{\text{all-cause}}_a(t)$) included in the demographic model are sourced from Office for National Statistics (ONS) estimates from 1981 until 2015. For years outside of the available data mortality rates are forecast using an age-stratified exponential model (\@ref(fig:mortality-england). This model was used as it constrains mortality rates above zero and decreases yearly changes in mortality rates over time. To model the uncertainty in the estimate of the annual number of deaths a normally distributed noise term was introduced with a standard deviation of 5%. In order to calculate the all-cause dynamic mortality rate ($\mu^{\text{all-cause}}_a(t)$), excluding deaths from, or related to, TB the following equation is used,

@@

Where $\mu_a(t)$ is constrained to be greater than or equal to zero.

```{r mortality-england, fig.scap = "3 year rolling average expected remaining lifespan stratified by age group in England from 2000 to 2014.", fig.cap = "3 year rolling average expected remaining lifespan stratified by age group in England from 2000 to 2014. Age is grouped into 5 year age groups from 0 to 49, from 50-69, and from 70 to 89. Those aged 90+ are excluded due to low quality data. The age groups used here represent those used in the model. Data from this figure was sourced from the ONS age-specific mortality rate estimates.", out.extra = ""}
knitr::include_graphics(file.path(resources_path, "resources", "figure", "mortality.png"))
```

>>>>

Non-UK born pulmonary cases was estimated using the ETS for each age-group included in the model from 2000 until 2015. Prior to 2000, incidence in the non-UK born are unavailable and the relationship to transmission in UK-born cases is unknown. To account for this, importation of non-UK born cases in the model begins in 1960 and then is scaled up through to 2000. As the form of this relationship is unknown the following functional form was used to scale cases based on those observed in 2000,

\begin{equation}
\text{Non UK-born cases (time=t)} = \frac{\text{exp}\left(\frac{t - 1960}{\text{ln}(2) * (-1) * \iota_{\text{scale}}}\right) - 1}{\text{exp}\left(\frac{2000 - 1960}{\text{ln}(2) * (-1) * \iota_{\text{scale}}}\right) - 1} * \text(Non UK born cases in 2000)
  (\#eq:non-uk-born-scale)
\end{equation}

@@

The probability of transmission can be defined as the probability that a single contact between an infectious active TB case and a susceptible individual will lead to TB infection. The probability of transmission ($\beta_a$) can be redefined in terms of effective contacts (the transmission probability times the contact rate; $C_{\text{eff}}$), historic effective contacts ($C^{\text{hist}}_{\text{eff}}$), actual average yearly total contacts ($C_{\text{actual}}$), the average period of time infectious ($\frac{1}{\nu_{\text{avg}}^{P}}$), and the average mortality rate ($\mu_{\text{avg}}$) as follows,

@@

Vynnycky et al. found that the effective contact rate for TB was approximately 22 in 1900 and fell to approximately 1 in 1990.[@Vynnycky1999] Incidence rates have increased since the early 1980s and it is unclear what impact this has had on the effective contact rate. I have assumed that the effective contact rate is normally distributed with a mean of 1 and a standard deviation of 0.5. For the historic effective contact rate I have assumed a uniform distribution with a lower bound of the current effective contact rate and an upper bound of 20. I have also assumed that the historic contact rate declines over time, starting in 1935, reducing to the current effective contact rate in 1980. The speed of this decay is set by defining a half-life ($C^{\text{half-life}}_{\text{eff}}$). The prior for the half-life is assumed to be normally distributed with a mean of 5 years and a standard deviation of 5 (truncated to be greater than 0). This prior is based on the observed trend in notifications. Additional age stratification of $\beta$ is explored by including modifiers for certain age-groups. The baseline scenario is that no modification is required, with variation explored for young adults (15-29; $\beta_{\text{young adult}}$) as a scenario. The priors for this modifier was assumed to be uniform, bounded by 0 and 10. The contact rate is estimated by averaging the total age-specific contact rates estimated from POLYMOD data (Section \@ref(contact-matrix)) on an annual basis.

@@

```{r tb-time-treat, fig.scap = "Distribution of time to treatment (days) from the date of reported symptom onset until the date started treatment for the UK born, stratified by age group and pulmonary/extra-pulmonary TB status in the ETS system for notifications between 2000 and 2012.", fig.cap = "Distribution of time to treatment (days) from the date of reported symptom onset until the date started treatment for the UK born, stratified by age group and pulmonary/extra-pulmonary TB status in the ETS system for notifications between 2000 and 2012. Age is stratified into three groups; children (0-14), adults (15-69) older adults (70-89). The time from symptom onset to starting treatment is shorter for cases with pulmonary TB cases across age groups, with younger cases starting treatment more rapidly than older cases. Vertical lines indicate the 2.5%, 25%, 50%, 75%, and 97.5% quantiles.", out.extra = ""}
knitr::include_graphics(file.path(resources_path, "resources", "figure", "time_to_treatment.png"))
```

@@

```{r tb-treat-time-succ, fig.scap = "Distribution of time to treatment completion in the UK born successfully treated (days), stratified by age group and pulmonary/extra-pulmonary TB status in the ETS for notifications between 2000 and 2012.", fig.cap = "Distribution of time to treatment completion in the UK born successfully treated (days), stratified by age group and pulmonary/extra-pulmonary TB status in the ETS for notifications between 2000 and 2012. Age is stratified into three groups;children (0-14), adults (15-69) older adults (70-89). There is little evidence that the time to successful treatment differs between pulmonary and extra-pulmonary cases only but older cases appear to have a high likelihood of longer treatment times. Vertical lines indicate the 2.5%, 25%, 50%, 75%, and 97.5% quantiles.", out.extra = ""}
knitr::include_graphics(file.path(resources_path, "resources", "figure", "time_cure.png"))
```

@@

In the model presented here unstratified social (nonphysical) contacts are used to generate an age-stratified contact matrix. There are several reasons for this. Firstly, stratifying by home, school, work, transport or leisure contacts, whilst initially appealing as doing so may lead to insights as to the nature of the type of contacts required for TB transmission, may lead to over-fitting without a strong apriori hypothesis. In high and medium burden countries it has been shown that within household transmission is not a major driver of overall transmission.[@Mathema2018] Until recently it has been thought that household transmission plays a more dominated role in low burden settings, such as England, which would indicate that home contacts should be considered. However, it has recently been found that 7.7% (1849/24,060) of cases in England between 2010 and 2012 lived in a household with another case.[@Lalor2017] The same study estimated that overall only 3.9% of cases were due to recent household transmission, and there was no evidence that cases within households were more likely to transmit within the household than outside of it. There is little evidence to suggest that school, work, transport or leisure contacts are more likely to transmit TB in England than any other contact. The choice of contact type is disease dependent; for TB it is likely that closer contacts result in a greater likelihood of transmission.[@Mathema2018] Unfortunately the physical contacts recorded in the POLYMOD survey represent a poor proxy to closeness of contacts as physical contact can be a little as a handshake and because TB is a respiratory disease physical contact is not required. For this reason physical contacts have not been further evaluated. Instead, the uncertainty in age-dependent transmission rates has been explored by allowing for scenarios in which the transmission probability varies between children and adults, between children, adults and older adults and in which it does not vary.

@@

As the POLYMOD contact data was collected using a survey there is likely to be measurement error and missing data for the number of contacts reported and the age that contacts were reported to be. Some participants also recorded contacts with an estimated age range rather than with a point estimate. In addition, as the survey had a relatively low sample size (1,011) in the UK, the estimated contact matrices contain considerable uncertainty. These considerations are often not considered in modelling studies but may introduce significant bias. Here the `socialmixr` R package^[`socialmixr`: https://github.com/sbfnk/socialmixr] is used to generate 1000 bootstrapped contact matrix samples using the following steps,

@@

Where $N_i$ is the number of people in age group $i$, $N_j$ is the number of people in age group $j$, $C_{ij}$ is the number of contacts between members of group $i$ with group $j$ and $C_{ji}$ is the number of contacts between members of group $j$ with group $i$. In the POLYMOD survey this relationship does not hold exactly due to random variation. A symmetric contact matrix ($C^{\prime}_{ij}$) can be derived by averaging the contacts between the $i$ and $j$ groups and the $j$ and $i$ groups for all age groups using the following equation,

\begin{equation} 
  C^{\prime}_{ij} = \frac{C_{ij}N_i + C_{ji}N_j}{N_i + N_j}
  (\#eq:symmfull)
\end{equation}

@@

```{r contact-tile-plot, out.width = "250px", fig.scap = "a.) Mean contacts (non-unique social contacts per year) and the b.) normalised standard deviation (\\%) of 1000 boostrapped samples of social contacts from the POLYMOD social contact survey using 5 year age groups up 49 years old and then a single group for 50-69 year olds.", fig.cap = "a.) Mean contacts (non-unique social contacts per year) and the b.) normalised standard deviation (\\%) of 1000 boostrapped samples of  social contacts from the POLYMOD social contact survey using 5 year age groups up 49 years old and then a single group for 50-69 year olds. Mixing is highly assortative by age with children and young adults representing the majority of contacts. There is also evidence of mixing between children and middle age adults with older children mixing with progressivly older adults. Contact rates in older adults are highly uncertain, with the most uncertainty in mixing between older adults and young children."}
knitr::include_graphics(file.path(resources_path, "resources", "figure", "contact_matrix.png"))
```

@@

An overview of the vacination model parameters can be found in Table \@ref(tab:disease-model) for parameters that impact the natural history of TB, Table  \@ref(tab:demo-model-params) for parameters that impact the population level distribution of BCG vaccination, and Table \@ref(tab:sources-tab) for an overview of the sources used to generate prior distributions. More detail is given in the following section.

@@

The effectiveness of the BCG vaccine is usually estimated using its effectiveness at reducing the incidence of active TB cases in a susceptible population. In the model outlined in this chapter the action of the BCG vaccine has been split into its main effect of reducing the rate of latent TB cases developing active disease and its secondary effect of reducing the likelihood of initial infection. There are few estimates of the effectiveness of the BCG vaccine at preventing active TB in cases that are already latently infected and where these estimates do exist they are not stratified by time since vaccination, or age at vaccination.[@Roy2014] The overall effectiveness ($\alpha^T_a$) of the BCG vaccine can be estimated from the combined effectiveness at preventing initial infection ($\chi_a$) and then preventing activation in latently infected individuals ($\alpha_a$) using the following equation,

@@

```{r bcg-eff-est}
kable(eff_bcg, booktabs = TRUE, escape = TRUE,
      caption.short = "Estimates of the effectiveness of the BCG vaccine at preventing active TB disease stratified by years since vaccination.", 
      caption = "Estimates of the effectiveness of the BCG vaccine at preventing active TB disease stratified by years since vaccination. For 0-9 years since vaccination estimates were derived using Poisson regression from the MRC BCG trial and for 10-29 years since vaccination estimates were extracted from a more recent case control cohort study in the UK born vaccinated at school-age. BCG effectiveness at preventing active TB disase used as prior distributions in the model.") %>% 
  kable_styling(latex_options = "hold_position")
```

@@

\begin{equation}
  \alpha^T_a = 1 - \text{exp}\left(\mathcal{N}(\text{ln}(RR_{t_v}, SE_{t_v})\right)
  (\#eq:rr-to-alpha)
\end{equation}

Where $RR_{t_v}$ is logged risk ratio and the $SE_{t_v}$ is the standard error of the logged Risk Ratio with both being dependent on the time since vaccination ($t_v$). The transformed values used as the prior distribution are detailed in Table \@ref(tab:disease-model).

@@

Roy et al. published a meta-analysis that estimated the effectiveness of the BCG vaccine at preventing initial infection in children.[@Roy2014] This has been used as the primary source for this parameter, with the assumption being made that the effectiveness is the same in adults as it is in children. This is reasonable to assume as there is little evidence the the overall effectiveness of the BCG vaccines reduces with the age it is given in England. Unfortunately the meta-analysis by Roy et al. did not include an estimate of the effectiveness of the BCG vaccine at preventing initial TB infection stratified by time since vaccination. This is problematic as there is a large amount of evidence that the overall effectiveness of the BCG vaccine wanes with time,[@Abubakar2013; @Mangtani2017] and if the protection from initial infection does not also reduce over time then as overall effectiveness decreases the contribution from the prevention of initial infection will increase. For this reason I have assumed that the protection from initial infection ($\chi_j$) reduces over time with the same functional form as for the overall effectiveness of BCG vaccination ($\alpha^T_j$). This relation can be formalised using the following equation,

@@

Age-stratified and UK birth stratified population estimates for England were estimated using the LFS (Section \@ref(tb-inc-rates-epi)). Figure \@ref(fig:age-strat-demo) indicates that the age distribution of the UK born population changed over the study period (2000 to 2015), with an increase in those in late middle age (45-49 years old) and older and a decrease in those in early middle age. The proportion of young adults and young children also increased. This may have impacted TB incidence as young adults are thought to be responsible for the majority of transmission. Data from the 1931 census was also used to estimate the population of England in 1931 stratified into the modeled age groups.

@@

The time varying, age-specific, noisy, all-cause mortality rates ($\mu^{\text{all-cause}}_a(t)$) included in the demographic model are sourced from Office for National Statistics (ONS) estimates from 1981 until 2015. For years outside of the available data mortality I forecast rates using an age-stratified exponential model (Figure \@ref(fig:mortality-england)). This model was used as it constrains mortality rates above zero and decreases yearly changes in mortality rates over time. To model the uncertainty in the estimate of the annual number of deaths a normally distributed noise term was introduced with a standard deviation of 5%. In order to calculate the all-cause dynamic mortality rate ($\mu_a(t)$), excluding deaths from, or related to, TB the following equation is used,

@@

Where $\mu_a(t)$ is constrained to be greater than or equal to zero, $\mu^{P}_a$ and $\mu^E_a$ are the age stratified deaths rates in pulmonary ($P_a$) and extra-pulmonary TB cases ($E_a$), and $N_a$ is the age stratified population.

```{r mortality-england, fig.scap = "3 year rolling average expected remaining lifespan stratified by age group in England from 2000 to 2014.", fig.cap = "3 year rolling average expected remaining lifespan stratified by age group in England from 2000 to 2014. Age is grouped into 5 year age groups from 0 to 49, from 50-69, and from 70 to 89. Those aged 90+ are excluded due to low quality data. The age groups used here represent those used in the model. Data from this figure was sourced from the ONS age-specific mortality rate estimates with projections based on an age-stratified exponential model.", out.extra = ""}
knitr::include_graphics(file.path(resources_path, "resources", "figure", "mortality.png"))
```
